Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-03 ms.tex
ref styles
aggghhhhhh not enough time

@@

One of the most remarkable results from the Kepler rotation period catalog was
the discovery of a bimodal period distribution among field stars by
\citet{mcquillan2013}, which is shown in Figure \ref{davenport}.
While a separate sequence of rapid rotation periods had been known in young
stellar clusters due to lower-mass stars settling on to the angular momentum
main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
detected from slower rotating M dwarfs (T$_\mathrm{eff} <$ 4000).
The bimodality separated M dwarfs into two populations, those with rotation
periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
and 20 days.
Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
period bimodality extended to include K dwarfs.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars, as shown in Figure
\ref{davenport} \citep{davenport2017}.
Two formation scenarios have been proposed to explain the observed period
bimodality.
The first scenario, initially proposed by \citet{mcquillan2013}, is the
rotation period distribution reflects the local star formation history, and
thus the bimodality represents a drop in the star formation rate around 600
Myr ago.
This model is supported by both the extension of the bimodality to earlier
spectral types by \citet{davenport2017}, and also the tentative detection that
the two rotation period populations have distinct proper motion distributions.
However, such a variation in the star formation rate on short timescales has
some tension with independent observational efforts to determine the local
star formation history.
While Color–Magnitude diagram inversions from Hipparcos have suggested a
similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
\citep{hernandez2000}, other studies find slower variations over several
Gyr \citep[\eg][]{cignoni2006}.
Using white dwarf cooling models to infer the local formation history
(“cosmochronology”) also supports higher star formation several Gyr ago, but
can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
sizes \citep{tremblay2014}.
The spatial extent of such coherent and localized variations in star formation
history is unknown.
The second scenario to explain this feature is that the period bimodality
occurs due to a previously unknown variation in the spin-down evolution for
low-mass stars.
In this model the star formation history would be continuous over the past
$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
period minima due to this unknown phase transition or feedback mechanism.
While this model is not predicted by angular momentum loss prescriptions,
rapid transitions in rotation period are observed for stars in young clusters.
Stars move quickly from the rapidly rotating “convective” sequence (periods of
$\sim<1$ day) to the “interface” sequence (periods of several days) during the
first few hundred Myr, with lower mass stars taking longer to make this
transition as they settle onto the main sequence \citep{barnes2003}.
Secondly, a gap in chromospheric activity levels for solar-type field stars
has also been observed \citep{vaughan1980}.
While this magnetic activity indicator smoothly varies with stellar ages over
long timescales, the gap indicates a rapid transition phase from “active” to
“inactive” is present within the first Gyr \citep{pace2009}.
Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
for stars older than the Sun, indicating a potential change in the magnetic
dynamo for slower rotators \citep{angus2015, vansaders2016}.
The TESS rotation period sample provides the ideal dataset to test these two
formation scenarios.
If the bimodality is due to an age distribution we would expect to only see
the feature locally, and that it could disappear at further distances or along
different lines of sight where small scale variations in the star formation
history are less apparent.
Coherent changes in the rotation period (or stellar age) distribution along
lines of sight might also reveal details of the spiral arm pattern speed near
the Sun.
The kinematic separation between the two rotation period populations would be
reinforced by supplemental measurements from the upcoming public Gaia data
releases.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, we would expect to find no little to no variation in
this feature as seen in Figure \ref{fig:davenport} with galactic position,
distance, or across the sky.

% \citet{McQuillan2014} demonstrated a striking discontinuity in the rotation
% periods of Kepler M dwarfs.
% They revealed a bimodal distribution of period, with a gap that depends on
% mass.
% Plenty of stars had rotation periods just shorter and just longer than this,
% but very few had rotation periods around this value.
% More recently, Davenport (2017) used Gaia data to elucidate this behaviour by
% re-examining the Kepler rotation periods reported in McQuillan et al. (2014)
% as a function of mass.
% Davenport (2017) eliminated contaminating giants identified using Gaia
% parallaxes, i.e. stars that are too bright to be
% dwarfs given their distance.
% These decontaminated data revealed that the period bimodality was not simply a
% feature of M dwarfs but extended to the G and K stars.
% Clearly then, this bimodal rotation period distribution is does not only
% affect M dwarfs, but may be a feature of all stars, including the Sun.

@@

% Intriguingly, the rotation gap is not constant across masses.
% It appears at shorter rotation periods for more massive stars and longer
% rotation periods for less massive stars.
% In fact, it closely follows a contour of constant age, an isochrone of 600
% Myr.
% Two possible scenarios can explain this phenomenon.
% The first is that stars undergo a transition in the behaviour of their
% magnetic dynamo.
% They switch from a fast to a slow rotating regime rapidly, spending little
% time rotating at an intermittent period.
% This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
% dwarfs.
% Is this a convergence bump?
% The rotation periods of stars are believed to have converged by around the age
% of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
% rotation periods and older clusters a narrow one).
% The age at which this transition occurs, around 600 Myr, is close to this.
% Could this be a coincidence or do rotation periods undergo a rotation period
% bump once they move stabilise onto a Skumanich spin-down regime?
% If this is the case, a study of this transition could greatly advance our
% understanding of the operation of stellar dynamos.

% The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
% (2017) could also be explained by two waves of star formation: a short one
% centered around a few hundred million years ago and a long one centered around
% a few billion years ago.
% TESS’s full view of the sky will allow us to establish the spatial extent of
% this rotation period gap and investigate the underlying cause.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, vansaders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{vansaders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation
period to the convective overturn time), its magnetic dynamo transitions into
a different state which drastically reduces (basically switches off) braking
efficiency.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
So, although this effect reduces the usefulness of gyrochronology for stars
more massive than the Sun, in practise, for Solar masses and below,
gyrochronology can still be considered an effective dating method.

Since Solar-type magnetic dynamos are generated through a combination of
convective motion and rotation, gyrochronology is thought to only be
applicable to stars with significant convective envelopes (T$_{\mathrm{eff}}
\lesssim$ 6250 K) that are not yet fully convective (3250 $\lesssim \mathrm{T}
_{\mathrm{eff}}$).

We intend to use the rotation periods of planet hosting stars to infer the
age-dependence of planet frequency. Kepler alone does not produce sufficient
stars with rotation periods to reliably characterise the time-dependent
frequency of exoplanets.
Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
transiting exoplanets in four young open clusters observed by the K2 mission.

@@

% \begin{thebibliography}
% \bibitem[Aigrain et al.(2016)]{aigrain2016} Aigrain, S., Parviainen,
%     H., \& Pope, B.~J.~S.\ 2016, \mnras, 459, 2408
% \bibitem[Angus et al.(2015)]{angus2015} Angus, R., Aigrain, S.,
% Foreman-Mackey, D., \& McQuillan, A.\ 2015, \mnras, 450, 1787
% \bibitem[Barnes(2003)]{barnes2003} Barnes, S.~A.\ 2003, \apj, 586,
% \bibitem[Barnes(2007)]{barnes2007} Barnes, S.~A.\ 2007, \apj, 669, 1167
% \bibitem[Cignoni et al.(2006)]{cignoni2006} Cignoni, M., Degl'Innocenti, S.,
% Prada Moroni, P.~G., \& Shore, S.~N.\ 2006, \aap, 459, 783 464
% \bibitem[Davenport(2017)]{davenport2017} Davenport, J.~R.~A.\ 2017, \apj, 835,
% 16
% \bibitem[Henderson et al.(2016)]{henderson2016} Henderson, C.~B.,
% Poleski, R., Penny, M., et al.\ 2016, \pasp, 128, 124401
% \bibitem[Hernandez et al.(2000)]{hernandez2000} Hernandez, X.,
% Valls-Gabaud, D., \& Gilmore, G.\ 2000, \mnras, 316, 605
% \bibitem[Luger et al.(2016)]{luger2016} Luger, R., Agol, E., Kruse, E., et
%     al.\ 2016, \aj, 152, 100
% \bibitem[Luger et al.(2017)]{luger2017} Luger, R., Kruse, E., Foreman-Mackey,
%     D., Agol, E., \& Saunders, N.\ 2017, arXiv:1702.05488
% \bibitem[Foreman-Mackey et al.(2017)]{foreman-mackey2017} Foreman-Mackey, D.,
%     Agol, E., Ambikasaran, S., \& Angus, R.\ 2017, arXiv:1703.09710
% \bibitem[McQuillan et al.(2013)]{mcquillan2013} McQuillan, A., Aigrain, S., \&
%     Mazeh, T.\ 2013, \mnras, 432, 1203
% \bibitem[McQuillan et al.(2014)]{Mcquillan2014} McQuillan, A., Mazeh, T., \&
%     Aigrain, S.\ 2014, \apjs, 211, 24
% \bibitem[Montet et al.(2017)]{montet2017} Montet, B.~T., Tovar, G., \&
%     Foreman-Mackey, D.\ 2017, arXiv:1705.07928
% \bibitem[Pace et al.(2009)]{pace2009} Pace, G., Melendez, J., Pasquini, L., et
%     al.\ 2009, \aap, 499, L9
% \bibitem[Skumanich(1972)]{skumanich1972} Skumanich, A.\ 1972, \apj, 171, 565
% \bibitem[Tremblay et al.(2014)]{tremblay2014} Tremblay, P.-E., Kalirai, J.~S.,
%     Soderblom, D.~R., Cignoni, M., \& Cummings, J.\ 2014, \apj, 791, 92
% \bibitem[Vaughan \& Preston(1980)]{vaughan1980} Vaughan, A.~H., \& Preston,
%     G.~W.\ 1980, \pasp, 92, 385
% \bibitem[Wang et al.(2016)]{wang2016} Wang, D., Hogg, D.~W., Foreman-Mackey,
%     D., \& Sch{\"o}lkopf, B.\ 2016, \pasp, 128, 094503
% \end{thebibliography}

\bibliographystyle{plainnat}
\bibliography{references}

>>>>

\bibliographystyle{aasjournal}

@@

One of the most remarkable results from the Kepler rotation period catalog was
the discovery of a bimodal period distribution among field stars by
\citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
While a separate sequence of rapid rotation periods had been known in young
stellar clusters due to lower-mass stars settling on to the angular momentum
main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
detected from slower rotating M dwarfs (T$_\mathrm{eff} <$ 4000).
The bimodality separated M dwarfs into two populations, those with rotation
periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
and 20 days.
Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
period bimodality extended to include K dwarfs.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars, as shown in Figure
\ref{fig:davenport} \citep{davenport2017}.
Two formation scenarios have been proposed to explain the observed period
bimodality.
The first scenario, initially proposed by \citet{mcquillan2013}, is the
rotation period distribution reflects the local star formation history, and
thus the bimodality represents a drop in the star formation rate around 600
Myr ago.
This model is supported by both the extension of the bimodality to earlier
spectral types by \citet{davenport2017}, and also the tentative detection that
the two rotation period populations have distinct proper motion distributions.
However, such a variation in the star formation rate on short timescales has
some tension with independent observational efforts to determine the local
star formation history.
While Color–Magnitude diagram inversions from Hipparcos have suggested a
similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
\citep{hernandez2000}, other studies find slower variations over several
Gyr \citep[\eg][]{cignoni2006}.
Using white dwarf cooling models to infer the local formation history
(“cosmochronology”) also supports higher star formation several Gyr ago, but
can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
sizes \citep{tremblay2014}.
The spatial extent of such coherent and localized variations in star formation
history is unknown.
The second scenario to explain this feature is that the period bimodality
occurs due to a previously unknown variation in the spin-down evolution for
low-mass stars.
In this model the star formation history would be continuous over the past
$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
period minima due to this unknown phase transition or feedback mechanism.
While this model is not predicted by angular momentum loss prescriptions,
rapid transitions in rotation period are observed for stars in young clusters.
Stars move quickly from the rapidly rotating “convective” sequence (periods of
$\sim<1$ day) to the “interface” sequence (periods of several days) during the
first few hundred Myr, with lower mass stars taking longer to make this
transition as they settle onto the main sequence \citep{barnes2003}.
Secondly, a gap in chromospheric activity levels for solar-type field stars
has also been observed \citep{vaughan1980}.
While this magnetic activity indicator smoothly varies with stellar ages over
long timescales, the gap indicates a rapid transition phase from “active” to
“inactive” is present within the first Gyr \citep{pace2009}.
Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
for stars older than the Sun, indicating a potential change in the magnetic
dynamo for slower rotators \citep{angus2015, van-saders2016}.

The TESS FFI and CTL targets will provide the ideal dataset to test these two
formation scenarios.
If the bimodality is due to an age distribution we would expect only to see
the feature locally; it could disappear at greater distances or along
different lines of sight where small scale variations in the star formation
history are less apparent.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, we would expect to find no little to no variation in
this feature as seen in Figure \ref{fig:davenport} with galactic position.

To map the rotation period distribution as a function of galactocentric
position we will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001}.
This will also allow us to remove contaminating subgiants and binary stars
from the sample, leaving only single main sequence stars.
The period distribution of G dwarfs is critical for this analysis since they
are observable at greater distances, allowing us to sample different star
formation histories.
With the April 2018 data release from Gaia (DR2) estimate that we will be able
to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3 kpc.
Unlike \Kepler, since \TESS\ is targeting the whole sky, it will be possible
to investigate the rotation period distribution as a function of
galactocentric position.
The size scale over which the Milky Way’s star formation history varies is
unknown however significant variations in star formation histories are seen in
Andromeda between 100 pc volumes, as well as large galaxy-wide trends
\citep[\eg][]{lewis2015}.

@@

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{van-saders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), magnetic braking efficiency effectively
drops to zero.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
Although this effect reduces the usefulness of gyrochronology for stars more
massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
effective dating method.
This will include a large number of \TESS\ planet candidates.

We intend to use the rotation periods of planet hosting stars to infer the
age-dependence of planet frequency.
Kepler alone has not yet produced a sufficient number of stars with rotation
periods to reliably characterise the time-dependent frequency of exoplanets.
Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
transiting exoplanets in four young open clusters observed by the K2 mission.

@@

\bibliography{references}
